---
title: "scTSS: a package to study transcription start site (TSS) based on 5' scRNA-seq
  data"
author: "Shiwei Fu, Wei Vivian Li"
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

scTSS is a bioinformatics package that leverages 5' single-cell RNA sequencing (scRNA-seq) data to predict transcription start sites (TSS) and investigate alternative TSS usage between different conditions or cell types. scTSS consists of two primary stages: TSS prediction and differential TSS usage (DU) testing.

During the TSS prediction stage, scTSS is capable of analyzing both single-end and paired-end data, yielding robust and accurate TSS predictions. In the DU testing stage, scTSS can perform DU tests across multiple conditions or cell types, providing users with reliable and informative results.

## Installation
`scTSS` is now available on GitHub. To download and load package please follow:

```{r  gh-installation, eval = FALSE}
devtools::install_github("https://github.com/ShweiSTAT/scTSS")
library(scTSS)
```

## Usage 
For more information, please use this [link](https://github.com/ShweiSTAT/scTSS/wiki/scTSS-Tutorial) to see the detailed tutorial. 

## Issues and communications
If you have any issues using this package, please post [here](https://github.com/ShweiSTAT/scTSS/issues). Any suggestions and comments are welcome! For suggestions and comments, please email to Shiwei Fu (shiwei.fu@email.ucr.edu) or Wei Vivian Li (weil@ucr.edu).
