---
title: "scTSS for studying transcription start sites using 5' scRNA-seq data"
author: "Shiwei Fu, Wei Vivian Li"
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

scTSS is a package that can leverage the 5’ single-cell RNA-seq data (both paired-end and single-end data) to study the transcription start site (TSS) at single-cell resolution. scTSS comprises of two main stages, TSS prediction and differential TSS usage (DU) test for TSSs across two or multiple conditions or cell types.

## Installation
`scTSS` is currently available on GitHub. To download and load package please follow:

```{r  gh-installation, eval = FALSE}
devtools::install_github("https://github.com/ShweiSTAT/scTSS")
library(scTSS)
```

## Usage 
For more information, please use this [link](https://github.com/ShweiSTAT/scTSS/wiki) to see the detailed tutorial. 

## Issues and communications
If you have any issues using this package, please post them [here](https://github.com/ShweiSTAT/scTSS/issues). Any suggestions and comments are welcome! For suggestions and comments, please contact Shiwei Fu (shiwei.fu@email.ucr.edu) or Wei Vivian Li (weil@ucr.edu).
